Conflict-Free Vertex Coloring Of Planar Graphs

### 2017-04-06 paper/paper.tex
Six-page draft so far...

The \emph{conflict-free coloring} problem is a relaxed variation of the vertex coloring problem. The conflict-free coloring problem does not aim to color every vertex such that it is not connected to a vertex of the same color. Rather, it aims to color \emph{enough} vertices such that for every vertex, it is connected to at least one uniquely colored vertex.

@@

A graph, denoted $G=(V,E)$, is an ordered pair of two sets: a set of vertices $V$ and a set of edges $E$. Each edge consists of a set of two unordered vertices from $V$. For example, $\{u, w\} \in E$ is an edge connecting vertices $u$ and $w$ where $u,w \in V$. Vertices are adjacent if they are connected by an edge. A simple graph is an undirected graph where each edge connects two different vertices and no two edges connect the same pair of vertices. All graphs used by the vertex coloring problem and conflict-free coloring problem will be undirected, simple graphs. An example graph that we will use later on is shown in figure \ref{fig:graph}.

@@

A path is a sequence of edges that connect a sequence of distinct vertices. A component of a graph $G$ is a subgraph $H$ where $H$ is connected and $H$ is not contained in any connected subgraph of $G$ that has more vertices or edges than $H$ has. An isolated path is a component that is itself a path.

A planar graph is a graph that can be drawn on the plane such that its edges only intersect at their endpoints. This drawing is referred to as a planar drawing. An outerplanar graph is a graph that has a planar drawing where every vertex belongs to the outer face of the drawing. \cite{bondy1976graph,west2001introduction}

@@

An example of a conflict-free coloring is shown in figure \ref{fig:cfcp-example}. We observe from figures \ref{fig:vcp-example} and \ref{fig:cfcp-example}, $\chi(G_1) = 3$ and $\chi_{CF}(G_1) = 2$. It is worth noting that all proper vertex colorings are also conflict-free colorings. This is because in a proper vertex coloring, every vertex is its own conflict-free neighbor \cite{abel2017three}.

@@

\subsection{Computational Complexity Theory}
\label{sec:complexitytheory}
A \emph{decision} problem, as mentioned earlier, is a problem that can be answered with a `yes' or a `no' \cite{sipser2006introduction}. A decision problem is said to be in the class \emph{P} if in the worst case, it can be solved with an algorithm that runs in polynomial time. A problem can be solved in polynomial time if an algorithm with input size $n$ can run in at most $n^k$ steps where $k$ is a constant that does not depend on $n$.

@@

There are certain problems, called NP-hard, that can be proven to be as hard as every problem in NP. A problem is NP-hard if it every problem in NP can be polynomially reduced to it. This means that if an NP-hard problem can be solved with a polynomial-time algorithm, then any problem in NP could be solved in polynomial time. A decision problem is said to be \emph{NP-complete} when it is both in NP and NP-hard.

@@

\section{Applications}
Before digging into methods for solving the conflict-free coloring problem, it is important to understand why conflict-free coloring is useful. It is useful to know what motivated the recent studies into conflict-free coloring.

@@

\begin{algorithm}
\caption{Iterated elimination of distance-3-sets} \label{alg:elimination}
\begin{algorithmic}[1]
\Input A simple, undirected graph $G$
\State $i \gets 1$
\State Remove all isolated paths from $G$
\While{$G$ is not empty}
	\State $D \gets \emptyset$
	\ForAll{components of $G$}
		\State Pick any vertex $v$
		\State $D \gets D \cup \{ v \}$
		\While{$\exists u$ at distance $\geq 3$ $\forall v \in D$}
			\State Pick $u$ at distance 3 from some vertex in $D$
			\State $D \gets D \cup \{ w \}$
		\EndWhile
		\ForAll{$u \in D$}
			\State Color $u$ with color $i$
		\EndFor
		\State $i \gets i + 1$
		\ForAll{$u \in D$}
			\State Remove $N[u]$ from $G$
		\EndFor
		\State Remove all isolated paths from G
	\EndFor
\EndWhile
\State Color all removed isolated paths using color $i$
\end{algorithmic}
\end{algorithm}

@@

It is wise to note that this algorithm does not necessariyminimize the number of colored vertices. It can be beneficial to minimize the number of colored vertices (less cost) in certain applications such as building wireless networks. The conflict-free domination number for $k=2$ of $G_3$, $\gamma_{CF}^2(G_3)$, is three even though we have four colored vertices in our example. This can be seen by coloring vertices $\{2, 8\}$ with one color and vertex 6 with another.

@@

\subsection{Bounds for Planar Graphs}
\label{sec:bounds}
Recall the map of the USA example given in the introduction. This is an example of a planar graph and illustrates a famous theorem associated with it. We present this theorem, called the four color theorem, in theorem \ref{thm:four}. It has been proven by multiple groups. The proof by computer that is widely accepted today can be found in \cite{robertson1997four}.

@@

Figure \ref{fig:four} looks at a planar graph that requires four colors in a proper vertex coloring. Based on theorem \ref{thm:four-cf}, we know that this planar graph can be conflict-free colored with only three colors.

\subsection{Minimizing Colored Vertices}
Abel et. al \cite{abel2017three} present specific results for 3-colorable graphs. They examine minimizing the number of colored vertices of 3-colorable graphs. They prove theorems \ref{thm:approximate-cfds} and \ref{thm:outerplanar-cfdn} by using a polynomial-time algorithm that transforms a dominating set $D$ of a planar graph $G$ into a conflict-free k-coloring of $G$ where only the vertices of $D$ are colored. This algorithm is defined in algorithm \ref{alg:cf-dominating}.

@@

Because the planar minimum dominating set problem is NP-hard \cite{garey2002computers}, we know that the conflict-free dominating set problem is NP-hard because we aim to find a minimum dominating set. It follows that the corresponding decision problem, the k-conflict-free dominating set problem, is NP-complete.

We get a PTAS for the conflict-free domination number by applying algorithm \ref{alg:cf-dominating} and finding the dominating set using the PTAS defined by Baker and Hill \cite{baker1994approximation}. As the input graph $G$ is planar, the resulting graph $G'$ is also planar because $G'$ is a minor of $G$. Thus, it can be colored with 4 colors by \ref{thm:four}. We then color $G$ with the colors of $G'$, leaving any vertex in $G$ but not in $G'$ uncolored. This leaves us with a conflict-free 4-coloring of $G$.

\begin{theorem} \label{thm:outerplanar-cfdn}
If a graph $G$ is outerplanar, then $\gamma_{CF}^k(G) = \gamma(G)$. Also, there exists a linear-time algorithm to compute $\gamma_{CF}^k(G)$.
\end{theorem}

\subsection{CF Coloring via Dominating Set}
Algorithm \ref{alg:elimination} presented in section \ref{sec:iterated} gives us an efficient heuristic for coloring general graphs including planar graphs. Algorithm \ref{alg:cf-dominating} gives us another way of coloring planar graphs, albeit likely less efficient.

@@

@@

Lastly, we color our original graph $G_3$ based on the proper vertex coloring we generated. We leave any nodes not in the graph $G$ uncolored when we color $G_3$. The coloring of $G_3$ is shown in figure \ref{fig:algorithm2-step4}. It is clear that this has produced a conflict-free coloring. Looking back at algorithm \ref{alg:elimination}, we had 4 colored vertices. Here, we have 3, and it is true that we have minimized colored vertices as $\gamma_{CF}^2(G_3) = 3$.

@@

>>>>

The \emph{conflict-free coloring} problem is a relaxed variation of the vertex coloring problem. The conflict-free coloring problem does not aim to color every vertex such that it is not connected to a vertex of the same color. Rather, it aims to color \emph{enough} vertices such that for every vertex, there is a unique color assigned to a vertex within its closed neighborhood.

@@

A graph, denoted $G=(V,E)$, is an ordered pair of two sets: a set of vertices $V$ and a set of edges $E$. Each edge consists of a set of two unordered vertices from $V$. For example, $\{u, w\} \in E$ is an edge connecting vertices $u$ and $w$ where $u,w \in V$. Vertices are adjacent if they are connected by an edge. A simple graph is an undirected graph where each edge connects two different vertices and no two edges connect the same pair of vertices. A planar graph is a graph that can be drawn on the plane such that its edges only intersect at their endpoints. This drawing is referred to as a planar drawing. All graphs used by the vertex coloring problem and conflict-free coloring problem will be undirected, simple graphs. An example graph that we will use later on is shown in figure \ref{fig:graph}.

@@

A \emph{path} is a sequence of edges that connect a sequence of distinct vertices. A component of a graph $G$ is a subgraph $H$ where $H$ is connected and $H$ is not contained in any connected subgraph of $G$ that has more vertices or edges than $H$ has. An isolated path is a component that is itself a path.

The distance of a path can be thought of the smallest number of edges needed to be crossed to get from one vertex to another vertex. For example, in figure \ref{fig:graph}, the distance from vertex 3 to vertex 4 would be 3 as it takes 3 edges to get there. A \emph{distance-3-set} is a set of vertices that all have exactly distance 3 from each other. An example of a distance-3-set from figure \ref{fig:graph} would be $\{3,4\}$. \cite{bondy1976graph,west2001introduction}

@@

An example of a conflict-free coloring is shown in figure \ref{fig:cfcp-example}. All uncolored vertices are in gray. It is a valid conflict-free coloring because every vertex has a unique color within its closed neighborhood. For example, $N(1) = \{1: uncolored,\ 2: uncolored,\ 5: red\}$. Thus, red is the unique color within the closed neighborhood of vertex 1. We observe from figures \ref{fig:vcp-example} and \ref{fig:cfcp-example}, $\chi(G_1) = 3$ and $\chi_{CF}(G_1) = 2$. It is worth noting that all proper vertex colorings are also conflict-free colorings. This is because in a proper vertex coloring, every vertex is its own conflict-free neighbor \cite{abel2017three}.

@@

\subsection{Computational Complexity Theory}
\label{sec:complexitytheory}
A \emph{decision} problem, as mentioned earlier, is a problem that can be answered with a `yes' or a `no' \cite{sipser2006introduction}. A problem can be solved in polynomial time if an algorithm with input size $n$ can run in at most $n^k$ steps where $k$ is a constant that does not depend on $n$. A decision problem is said to be in the class \emph{P} if in the worst case, it can be solved with an algorithm that runs in polynomial time.

@@

There are certain problems, called NP-hard, that can be proven to be as hard as every problem in NP. If problem $Y$ can be transformed into problem $X$ by a polynomial-time algorithm, it is said that $Y$ can be polynomially reduced to $X$. A problem is NP-hard if it every problem in NP can be polynomially reduced to it. This means that if an NP-hard problem can be solved with a polynomial-time algorithm, then any problem in NP could be solved in polynomial time. A decision problem is said to be \emph{NP-complete} when it is both in NP and NP-hard.

@@

\section{Applications}
Before digging into methods for solving the CFCP, it is important to understand what motivated the recent studies into conflict-free coloring.

@@

\begin{algorithm}
\caption{Iterated elimination of distance-3-sets} \label{alg:elimination}
\begin{algorithmic}[1]
\Input A simple, undirected graph $G$
\State $i \gets 1$
\State Remove all isolated paths from $G$
\While{$G$ is not empty}
	\State $D \gets \emptyset$
	\ForAll{components of $G$}
		\State Pick any vertex $v$
		\State $D \gets D \cup \{ v \}$
		\While{$\exists u$ at distance $\geq 3$ $\forall v \in D$}
			\State Pick $u$ at distance 3 from some vertex in $D$
			\State $D \gets D \cup \{ w \}$
		\EndWhile
		\ForAll{$u \in D$}
			\State Color $u$ with color $i$
		\EndFor
		\State $i \gets i + 1$
		\ForAll{$u \in D$}
			\State Remove $N(u)$ from $G$
		\EndFor
		\State Remove all isolated paths from G
	\EndFor
\EndWhile
\State Color all removed isolated paths using color $i$
\end{algorithmic}
\end{algorithm}

@@

It is wise to note that this algorithm does not necessarily minimize the number of colored vertices. It can be beneficial to minimize the number of colored vertices (less cost) in certain applications such as building wireless networks. The conflict-free domination number for $k=2$ of $G_3$, $\gamma_{CF}^2(G_3)$, is three even though we have four colored vertices in our example. This can be seen by coloring vertices $\{2, 8\}$ with one color and vertex 6 with another.

@@

\subsection{Bounds for Planar Graphs}
\label{sec:bounds}
Recall the map of the USA example given in the introduction. This is an example of a planar graph and illustrates a famous theorem associated with it. We present this theorem, called the four color theorem, in theorem \ref{thm:four}. The proof by computer that is widely accepted today can be found in \cite{robertson1997four}.

@@

Figure \ref{fig:four} looks at a planar graph that requires four colors in a proper vertex coloring. Based on theorem \ref{thm:four-cf}, we know that this planar graph can be conflict-free colored with only three colors. This is shown in figure \ref{fig:four} as well.

\subsection{CF Coloring via Dominating Set}
\label{sec:coloring-dom}
Algorithm \ref{alg:elimination} presented in section \ref{sec:iterated} gives us an efficient heuristic for coloring general graphs including planar graphs. Algorithm \ref{alg:cf-dominating}, called conflict-free via dominating set, gives us another way of coloring planar graphs, albeit likely less efficient.

@@

@@

Lastly, we color our original graph $G_3$ based on the proper vertex coloring we generated. We leave any nodes not in the graph $G$ uncolored when we color $G_3$. The coloring of $G_3$ is shown in figure \ref{fig:algorithm2-step4}. It is clear that this has produced a conflict-free coloring.

@@

Take note that this algorithm has produced fewer colored vertices than algorithm \ref{alg:elimination}. We had 4 colored vertices based on that algorithm where here we have 3. This algorithm helps us prove theorems based on minimizing the number of colored vertices.

@@

\subsection{Minimizing Colored Vertices}
Abel et. al \cite{abel2017three} present results on minimizing colored vertices when conflict-free coloring planar graphs. They propose and prove theorem \ref{thm:approximate-cfds} by using the polynomial-time algorithm described in algorithm \ref{alg:cf-dominating}.

Because the planar minimum dominating set problem is NP-hard \cite{garey2002computers}, we know that the conflict-free dominating set problem is NP-hard because we aim to find a dominating set. It follows that the corresponding decision problem, the k-conflict-free dominating set problem, is NP-complete.

We get a PTAS for the conflict-free domination number by applying algorithm \ref{alg:cf-dominating} and finding the dominating set using the PTAS defined by Baker and Hill \cite{baker1994approximation}. As the input graph $G$ is planar, the resulting graph $G'$ is also planar because $G'$ is a minor of $G$. Thus, it can be colored with 4 colors or less by \ref{thm:four}. We then color $G$ with the colors of $G'$, leaving any vertex in $G$ but not in $G'$ uncolored. This leaves us with a conflict-free coloring of $G$.
